Stage-Splitting a Modal Language

### 2014-10-07 paper/splitting.tex
major work on 4.0 - 4.3

As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$, 
which respectively split terms that type $@ \bbone$ and $@ \bbtwo$. 
Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.

\subsection{Partial Value Splitting}

\ur{Reduce commentary.  Two sentences connected with an ``and.'' Split
  sentences.}

We now consider the problem of splitting stage~\bbone\ partial values.
This serves as a simple didactic stepping-stone to the full formulation, 
and we will see later that it is a necessary component of the correctness theorem.

Recall that, along with a future environment, partial values are the output of the partial evaluation of stage \bbone\ terms.
In addition to base types, tuples, and injections, partial values may also contain references into the future environment.
In this way, partial values express not only stage \bbone\ content, but stage \bbtwo\ content as well.
The goal of partial value splitting is to take a partial value $v$ and from it create two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to first stage code inspecting this value, 
and $v_2$ contains only the information from $v$ that is likewise relevant to second stage code.
We call this action {\em masking}.

\ur{Next sentence is unnecessary commentary. It also sounds of
  questinoable accuracy, not clear to me that there are many ways to
  do this.}  There are many reasonable ways to achieve this goal, and
we choose one that preserves as much structure as possible from $v$.
Formally, partial value splitting comprises a single judgment, $v
\vsplito [v_1,v_2]$, the rules for which are given in
\cref{fig:valSplit}.  It essentially acts distributively at product,
sum, and recursive types, doing little more than recuring into
subvalues and duplicating the structure to both results.  At base
types, partial value splitting assigns all content ({\em i.e.} the
integer or boolean) to stage \bbone, and uses a nullary tuple
placeholder for stage \bbtwo.  Conversely at $\fut$ types, splitting
assigns all content ({\em i.e.} the environment reference) to stage
\bbtwo, and uses a nullary tuple for stage \bbone.

@@

\ur{Drop, ``the most we can say.''  Be direct and explain.}
Functions are inherently more complicated.  Since the body of a
function is an open term, the partial value splitting rule for
functions must rely on general expression splitting, which is covered
momentarily.  The most we can say here is a function splits into two
other functions, whose bodies correspond to the stage \bbone\ and
stage \bbtwo\ components of the original function's body.

\subsection{Expression Splitting}

We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
These are respectively called the {\em precomputation} and {\em residual}.
\footnote{We also used the term {\em residual} for the result of
  partial evaluation.  Where necessary, we will distinguish between
  these concepts by using the terms {\em splitting-residual} and {\em
    evaluation-residual}.}  \ur{The reuse of residual is bothersome,
  we should use different terminology, 1-value and 2-value perhaps.}

\ur{We need a ``next'' I believe.  All our examples should start at
  stage-1 for consistency.}
For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
an action we call {\em lifting}.

\ur{I have no idea what the next paragraph is trying to say.}
Note that lifting and masking are different, despite both being within the purview of splitting.
They particularly differ in how the resulting stage \bbone\ parts are used.
The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
and it's accessible to whatever surrounding piece of code refers to the value being split.
In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
and is not accessible externally.

The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
The stage \bbone\ result of masking is \verb|((),3*4)|, 
and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
This is called the {\em combined result} for stage \bbone.
\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}
The splitting residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
For our example, it is ``\verb|fn L => (L<5,())|".

@@

\ur{``for both'' below is ambiguous.}
The splitting algorithm comprises two judgments, $\splitonesym$ and $\splittwosym$,
which are in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
For both, the input is a term in \lang, and the output is two terms in \langmono, an
unstaged language.  The grammar for \langmono\ is given in \ref{fig:monoGrammar}.  It has no
staging features (we say that implicitly it has only one stage), and it is untyped.  
Modulo these differences, \langmono\ has all of the same features as \lang, 
although this is a matter of taste
\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.
\ur{Drop commentary about taste and such.  Explain that we could have
  used lambda calculus as target but this is more direct.}

\ur{Drop ``specifically''.}
Specifically, the $\splitonesym$ judgment sends a stage \bbone\ term ($e$) to a combined term ($c$) and residual ($l.r$),
while $\splittwosym$ sends a stage \bbtwo\ term ($e$) to a precomputation ($p$) and a residual ($l.r$).
For concision, we represent residuals as a term open on a single variable, rather than as functions.
For both judgments, we use a context ($\Gamma$) to keep track of the
open variables of $e$.
\ur{Case (pl./si.) mismatch in ``we represent ...'' sentence above.}

@@

\ur{Explain the rules.  This is the most important part of the
  paper. Work needed here.}
The rules\footnote{The rules are written using patterns, including the open variable of the residual.} 
for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the residual\footnote{The residual of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

The rules of stage \bbtwo\ splitting are given in \ref{fig:termSplitTwo}.  
Every rule works by bundling the precomputations of the constituent parts, and then unbundling them with a pattern.
The rules of stage \bbone\ splitting are given in \ref{fig:termSplitOne}.

\subsection {Properties}

We'll now call out a few properties to notice in the splitting rules.

\subsubsection {Expression Splitting at Values}

Note that general expression splitting always produces an effectively trivial precomputation when applied to partial values,
where effectively trivial means that the precomputation is either a unit value, or a tuple or injection of an effectively trivial values.

Partial value splitting is more than just a restriction of general expression splitting 
precisely because it observes this property it at the structural level, doing away with the precomputation entirely.
This makes sense intuitively, since a precomputation would imply that there is some computation left to do, 
which is antithetical to the notion of values.

\subsubsection {Speculation}

Notice in the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the precomputation is lifted out from within branches.
This is the manifestation of the speculation behavior from the semantics.

\subsubsection {Stage \bbone\ Divergence}
Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject precomputation into a sum type.
Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.

\subsubsection {Stage \bbone\ Functions}

The trick when splitting stage \bbone\ functions is that the contents themselves may be multi-stage.
We handle this by splitting them into two functions:
one in the immediate result which handles all the stage \bbone\ content of the original,
and one in the residual which handles all of the stage \bbtwo\ content.
Note that stage \bbone\ $\lambda$-expressions themselves have only a trivial precomputation.

>>>>

%As with the dynamics, our stage-splitting algorithm for expressions in \lang\
%takes the form of two mutually-dependent judgments, $\splitonesym$ and
%$\splittwosym$, which respectively split terms that type $@ \bbone$ and $@
%\bbtwo$.  Before diving into the full generality, 
As discussed in the introduction, a stage splitter must take a term in \lang\
and produce two

\TODO waiting on how the intro defines splitting

In this section, we define splitting for partial values, then extend it to all
terms, and prove correctness of this algorithm.

\subsection{Partial Values}

Partial values, along with future environments, are produced by partially
evaluating stage \bbone\ terms. Although partial values have no more stage
\bbone\ computation remaining, they may contain variables bound by the future
environment, and thus, may require stage \bbtwo\ computation.

Splitting a partial value $v$ creates two single-stage values $v_1$ and $v_2$,
where $v_w$ contains only the information required for stage $w$ evaluation of
$v$. Splitting partial values is implemented by the judgment $v \vsplito
[v_1,v_2]$ defined in \ref{fig:valSplit}. 
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
exposing only the relevant structure of a term is called \emph{masking}.

@@

Splitting a function is more complex, because function bodies are open terms,
and so are split by the judgments described in the next subsection. Essentially,
functions will split into two functions, whose bodies correspond to the stage
\bbone\ and stage \bbtwo\ components of the original function body.

\subsection{General Splitting}

\ur{The reuse of residual is bothersome, we should use different terminology,
1-value and 2-value perhaps.}

Splitting arbitrary terms in \lang\ again results in two terms containing only
the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
called, respectively, the {\em precomputation} and {\em residual}.%
\footnote{We also used the term {\em residual} for the result of
partial evaluation.  Where necessary, we will distinguish between
these concepts by using the terms {\em splitting-residual} and {\em
evaluation-residual}.}
%
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a residual
\verb|fn (L1,L2) => L1 < L2-5|, with the property that the residual applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{lifting}.

Lifting and masking are two necessary ingredients for ensuring that splitting
observes the stages of \lang. When splitting a partial value, the stage \bbone\
result must \emph{mask} the stage \bbtwo\ computations in order to be
single-staged. When splitting a stage \bbtwo\ term, the precomputation must
\emph{lift} out all embedded stage \bbone\ computations, so they can be
performed first, even though their results are only used in stage \bbtwo.
%and it's accessible to whatever surrounding piece of code refers to the value being split.
%is not accessible externally.

Splitting a stage \bbone\ term, in contrast, requires both masking and splitting.
Consider the term 
\begin{lstlisting}
1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
\end{lstlisting}
Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
must occur at stage \bbone, the former is needed by stage \bbone, while the
latter is needed by stage \bbtwo.

The precomputation for this term is a \emph{combined result} which is
the pair of these components: \verb|(((),3*4),1+2)|. The splitting residual is
likewise the pair of the stage \bbtwo\ results of lifting and masking:
\verb|fn L => (L<5,())|.
%\footnote{In this example, the combined result is a 2-tuple, but more generally
%it only must be a term which reduces to a 2-tuple.}

@@

%in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
Splitting is implemented by the judgments $\splitonesym$ and $\splittwosym$,
which take stage \bbone\ and stage \bbtwo\ terms respectively; these judgments
are defined in \cref{fig:termSplitOne,fig:termSplitTwo} inductively on the
structure of the term. Both judgments produce two terms in an unstaged language,
\langmono, whose grammar is given in \ref{fig:monoGrammar}. This language is
untyped and contains no staging features, but otherwise has the same term
constructors as \lang.
%\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.

The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and residual $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
precomputation $p$ and a residual $l.r$. In both cases, the term $e$ is open on
variables in the context $\Gamma$. We write the rules using pattern notation
when appropriate.
%For concision, we represent residuals as a term open on a single variable,
%rather than as functions.

@@

\ur{Explain the rules.  This is the most important part of the
  paper. Work needed here.}
Every $\splittwosym$ rule works by bundling the precomputations of the constituent parts, and
then unbundling them with a pattern.
The rules for splitting $\next$, $\prev$, and $\pause$ are given in \ref{fig:termSplitOne}.
The rule for $\next$ simply tuples up the precomputation of its subexpression with a trivial immediate result,
while the rule for $\prev$ projects the combined result of its subexpression to just the precomputation.
The $\pause$ rule treats the entire combined result of its subexpression as a precomputation, 
and projects out the integer result in the residual.%
\footnote{The residual of an integer expression is usually trivial, 
but we have to include it here for termination purposes.}

There are some particularly notable features of this definition of term
splitting.

\paragraph {Expression Splitting at Values}

When applied to partial values, term splitting always produces a trivial
precomputation, in the sense that the output is a nested sequence, possibly
empty, of tuples and injections of \texttt{()}. The intuition is that a partial
value has no remaining stage \bbone\ computation, and hence no precomputation.

%Partial value splitting is more than just a restriction of general expression
%splitting precisely because it observes this property it at the structural
%level, doing away with the precomputation entirely.

\paragraph {Speculation}

In the stage \bbtwo\ rules for {\tt if}s, {\tt case}s, and functions, the
precomputation is lifted out from within branches. This is the manifestation of
the speculation behavior from the semantics.

\paragraph {Stage \bbone\ Divergence}

In the splitting rules for stage \bbone\ {\tt if} and {\tt case} expressions, we
have enough information at stage \bbone\ to know what branch to take, so there's
no need to speculate. Instead, we evaluate the stage \bbone\ portion of the
taken branch only, and inject the precomputation into a sum type. In the
residual, we case on that injection, and resume the stage \bbtwo\ portion of the
correct branch.

\paragraph {Stage \bbone\ Functions}

The body of a stage \bbone\ function may contain stage \bbtwo\ computation. We
handle this by splitting the function into two functions: one in the immediate
result which handles all the stage \bbone\ content of the original, and one in
the residual which handles all of the stage \bbtwo\ content. Note that stage
\bbone\ lambdas are partial values, and so have only a trivial precomputation.
